Why there are more species in several small patches vs few large patches:

### 2024-02-02 RunSims.Rmd
Update to figures and ms

# Shannon-Diversity  
#
mdl1 %>%  
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birth_rate_birds), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ birth_rate_birds, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Number of species") +
  theme(
    legend.position = c(0.9,0.8)
    )

@@

mdl <- sloss_sim %>% filter(deforestation_type == "regular", habitat_patch_size != 199)
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,birds_behavior) %>% summarise(across(birds_at_deforestation:count_birds,list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)

@@

# Shannon-Diversity  
#
mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birds_behavior), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ birds_behavior, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.9,0.8)
    )

>>>>

@@

mdl <- sloss_sim %>% filter(deforestation_type == "regular", habitat_patch_size != 199)
names(mdl)
#
# summarize data 
#
mdl1 <- mdl %>% mutate(habitat_patch_size = habitat_patch_size+1) %>%
                         group_by(birds_dispersal_distance, habitat_patch_size,birth_rate_birds,birds_behavior) %>% summarise(across(birds_at_deforestation:count_birds,list(mean=mean, sd=sd, ql=~quantile(.x,probs=c((1-0.95)/2)), qh=~quantile(.x,probs=c(1-(1-0.95)/2)) ))) #  %>% mutate(deltaPop = (count_birds - birds_at_deforestation )/ birds_at_deforestation  )
names(mdl1)
behavior_order <- c("NoSelection", "Hierarchical", "BirthSelection")
behavior_labels <- c("Neutral", "Hierarchical", "Habitat Selection")

mdl1 <- mdl1 %>% mutate(birds_behavior= factor(birds_behavior, levels = behavior_order, labels = behavior_labels))

@@

# Shannon-Diversity  
#
mdl1 %>%  filter(birth_rate_birds == 2) %>% 
  ggplot(aes(x = habitat_patch_size, y = calc_shannon_diversity_mean, color = factor(birds_dispersal_distance))) + 
  geom_line(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = .5) +
  geom_point(aes(group = interaction(birds_dispersal_distance, birth_rate_birds)),  size = 1.2) +
  geom_ribbon(aes(ymin = calc_shannon_diversity_ql, ymax = calc_shannon_diversity_qh, group = interaction(birds_dispersal_distance, birds_behavior), fill = factor(birds_dispersal_distance)), alpha = 0.3,colour=NA, show.legend = FALSE) +
  scale_color_viridis_d() + scale_fill_viridis_d() +
  facet_wrap(~ birds_behavior, ncol = 4) + 
  labs(color = "Dispersal\n distance") +
  theme_bw() + xlab("Habitat Patch Size") + ylab("Shannon Diversity") +
  theme(
    legend.position = c(0.85,0.18)
    )

### 2024-02-02 ms_SLOSS.Rmd
More languaje updates to the ms
Update to figures and ms

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. Still, the policies of many conservation agencies are prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), besides empirical support favouring SS. However, to date, most of the theories regarding this issue were based on single-species models. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters, when there is a reproduction event they send the propagule to a surrounding patch blindly, if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to habitat loss [@Young2016;@Chase2020;@Newbold2015]. Moreover, habitat fragmentation occurs at the same time as habitat loss, and the effects of both processes are difficult to disentangle [@Haddad2015]. Landscape connectivity and organism mobility are critical features for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with a lack of knowledge of most of the world’s biodiversity. Early recommendations claimed to follow ecological theory, in particular island biogeography theory to derive simple thumb rules decisions to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but only with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this has led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

The SLOSS debate has been an important academic and practical discussion in conservation biology. In practice, a series of decision-maker targeted recommendations were proposed and remain rather uncritically in several conservation instances (quotes).
While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss [@Keil2015;@Hodgson2011], the proposition that single large reserves retain more species than several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms (reviewed in  @Fahrig2017). Moreover, empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020], add more. The actual processes driving to it remain elusive and it is probably multi-causal (quotes). In that scenario, model simulations could produce useful predictions targeting the relevant variables to better understand the processes leading to non-intuitive patterns.

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics. Recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

We considered only one amount of habitat destruction in combination with different community dynamics, neutral and competitive, taking into account healthy and impoverished communities with different dispersal related traits, and different patterns of habitat loss.

This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. In particular, we simulate species retention after different fragmentation scenarios of the same overall habitat loss, therefore accounting for the interacting effects of habitat loss and fragmentation.

@@

We developed a spatially explicit multi-species model to explore the relationship between habitat loss and species preservation. In this model space is discretized into a grid and only one individual of one species can occupy a particular position. Each position represents an area fixed by the investigator to mimic the real system. Time is continuous, so the update of the model is asynchronous. Sites are selected at random and to perform one complete time interval $J$ sites have to be updated, where $J$ is the size of the grid.

We use absorving boundary conditions, which means that if an individual try to disperse outside the grid it dies. The size of the community is given by *J = dimX* x *dimY*, where *dimX* and *dimY* are the dimensions of the grid. Thus *J* is the maximum number of individuals in the simulated area. As in a classical neutral model, there is a metacommunity, i.e. a regional species pool from where species migrate [@Hubbell2001]. All individuals have the same parameters and each species is assigned an indicator number.

@@

The classical neutral variation where propagules (e.g., seeds) disperse passively and cannot actively choose their establishment site. During reproduction events, individuals disperse their propagules blindly to surrounding patches. If a patch is already occupied, the propagule fails to establish, resulting in its death. This variation captures the random dispersal and colonization processes typical of many plant species.

The neutral variation with habitat selection where individuals can actively select suitable habitat patches. In this variation, propagules can choose unoccupied patches, ensuring their survival and successful establishment. This modification reflects the influence of habitat availability and selection on species colonization and persistence. By incorporating habitat selection, this variation better captures the ecological processes and behaviors of species that are capable of actively seeking out suitable habitats for reproduction and survival.

The competitive hierarchical variation extends the neutral variation with an additional layer of complexity by incorporating competitive interactions among species. Each species has an indicator number that determines its competitive dominance level, and more dominant species can replace less dominant ones at a specified replacement rate. This variation captures the influence of species interactions and competitive exclusion on species composition and abundance. It is particularly relevant for understanding communities where competitive interactions play a significant role in shaping species distributions and community structure.

@@

    Regular degradation: Habitat is removed so that habitable patches form a regular pattern of squares with side given by an habitat-size parameter. As the percentage of degraded habitat is constant varying the habitat-size parameter results in either a few large habitat patches (SL) or several small (SS).

    Random block degradation: Habitat square patches of a fixed size (habitat-size parameter) are located at random throughout the landscape. The sizes are identical to those used for regular degradation simulations, but it results in habitat patches that have a minimum size corresponding to the habitat-size  but a maximum size that is bigger because habitat patches can superpose between them producing a larger patch. Then, I characterize this with the maximum habitat patch size.

(2) With $migration-rate$ an individual of a species $i$ can migrate from the metacommunity. A random site on the border of the grid is selected and if it is empty and habitable the site is occupied.

@@

    where $d(x)$ is the probability that an individual disperses a distance $x$ from the parent. In all cases we used $x_{min} = 1$. We use as parameter the mean $dispersal-distance$.

@@

    b. Neutral with Habitat Selection:  Propagules in this version selectively target unoccupied patches for reproduction within the distance given by the dispersal kernel, ensuring their survival.

@@

For all simulations we used a metacommunity of 100 species. The simulations commence with the colonization of an empty habitat area. After 600 time steps fixed amount of 60% habitat is subsequently subjected to destruction, employing regular or random block degradation to mimic habitat fragmentation due to habitat loss. <!-- why we don't use completely random degradation patterns -->

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist, but in spatially explicit models the value needs to be higher, around 1.6. Thus we used a minimum $\lambda$ of 1.7, we also fixed the $migration-rate$ at 0.0001. We simulate each model variant using the parameters shown in Table 1. For the neutral classical version only, we also simulate random block habitat patterns. For each combination of parameters we ran 30 simulations up to 1500 time steps and recorded the number of species, the Shannon Diversity index [@Magurran2013] and the total number of individuals. For some representative combinations of parameters we take snapshots of the model output showing the typical dynamics on time of the 3 recorded quantities and the rank abundance species distribution [@Newman2005; @May1975]

@@

The model was developed in Netlogo 6.3.0 and model source code is available at Github <https://github.com/lsaravia/DynamicForestExtinction>, the Overview, Design concepts, Details (ODD), protocol for describing individual-based models is included within the model source code [@Grimm2010]. We performed the simulations and analysis using R version 4.3.1 [@RCoreTeam2023], the R source code to perform simulations and analysis is located at <https://github.com/EcoComplex/multiSLOSS>.

## Results

In our neutral model simulations with a migration rate of \(0.0001\) and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the \(\lambda\) parameter (Figure 1). Species richness increased with \(\lambda\), with the highest richness at smaller patch sizes for higher \(\lambda\) values. Conversely, a \(\lambda\) of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. The interplay of dispersal distances, habitat sizes, and \(\lambda\) impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low \(\lambda\) and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

Our research contributes to the ongoing Single Large vs Several Small (SLOSS) debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under similar conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) and dispersal relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and patch sizes are above a certain threshold to prevent local extinctions, aligning with Rybicki's [-@Rybicki2020] findings.

In neutral models one species by chance tends to dominate a patch [@Hubbell2001], so in fragmented landscapes, once the species increments its density is very improbable that another species replaces it so with more patches the probability of having more species is augmented. This effect is enhanced when individuals can select the habitat, because of the increased survival of the propagules, thus fragmentation has a positive effect on the number of species. 
For competitive communities, there is a higher probability that different species dominate different smaller patches, compared with a bigger patch. Thus, again fragmentation has a positive effect on the number of species, this effect is called stochastic divergence [@Riva2023]. In smaller patches there is an increase in extinction risk because they support smaller populations [@Laurance2002;@Shafer1995], however as patches become smaller they also become more numerous, reducing the probability of extinction over the entire set of patches due to spreading the risk effect [@Crowley1977]. The combination of these two effects, the stochastic divergence and the risk effect, explain the highest number of species with patch size over the threshold of species survival.

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not included in the model, environmental heterogeneity, also the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

@@

The novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat, and then after the number of species is stabilized the habitat is fragmented. This could be a more realistic scenario than if all species are present with initially random patterns at an already fragmented habitat [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation.

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches, such as isolated trees, in maintaining biodiversity and landscape connectivity [@Yan2021; @Manning2006; @Shafer1995]. Our results thus challenge the conventional preference for larger reserves in conservation planning and support the paradigm shift in conservation towards considering sets of many small patches generally having higher biodiversity than sets of few large patches [@Riva2023].

In conclusion, our study adds valuable perspectives to the SLOSS debate, illustrating the complexities of habitat fragmentation and species conservation. While our model has inherent limitations, it provides a useful framework for understanding key ecological processes. Future research should aim to extend these findings to more diverse ecological settings and incorporate additional factors influencing biodiversity in fragmented landscapes.

@@

>>>>

As global biodiversity is rapidly declining due to habitat loss, it is important to determine how to protect it. The policies of many conservation agencies are still prioritizing the preservation of a single large habitat area (SL) vs. several small areas of the same size (SS), despite empirical support favouring SS. However, to date, most theories regarding this issue were based on single-species models. This work aims to explore spatial multi-species models to verify the SS > SL pattern and determine the causes of it. We used three types of models: a classical neutral, neutral with habitat selection, and a competitive hierarchical model.  All models have three parameters, namely extinction, reproduction, and dispersal distance. In the classical neutral model, species are equivalent and have the same parameters; when there is a reproduction event they send the propagule to a surrounding patch blindly, and if the patch is already occupied the propagule dies. In the neutral-with-habitat selection model, the propagules select an empty patch so their survival is ensured. In the hierarchical model, species have competitive dominance and more dominant ones can replace the less dominant with a replacement rate parameter. We use 100 species and start simulating the colonization of an empty area then we destroy a fixed amount of habitat using different regular patch sizes. The results for all models are almost always that SS retains more species than SL. The extinction of species in patches depends on the quotient between reproduction and extinction rates in relation to patch size. The only case when SL > SS is when dispersal is insufficient to connect patches, and patch size is smaller than the minimum to prevent extinction. If patch size is above the critical size to maintain local populations the SS > SL pattern is due to the sampling effect of aggregated species distribution combined with the reduction of dispersal ability of species. High dispersal rates produce homogenization of species distribution and a reduction of species richness and this is why habitat fragmentation by habitat loss retains more species when patches are smaller. The model predicts that if a set of species have more dispersal capabilities the SS effect would be lower than for species with less dispersal.

@@

Global biodiversity is rapidly declining due to, among other factors, habitat loss [@Young2016;@Chase2020;@Newbold2015]. As habitat fragmentation occurs at the same time as habitat loss, the effects of both processes are difficult to disentangle [@Haddad2015]. Landscape connectivity and organism mobility are critical features for population persistence [@Taylor1993]. Starting from a landmark paper [@Diamond1975], conservation biologists engaged in the challenging task of recommending protection measures given the environmental crisis coupled with a lack of knowledge of most of the world’s biodiversity and limited funding for conservation. In that context, early recommendations claiming to follow ecological theory derived simple thumb rules decisions to better design protected areas. However, as noted rapidly, species/area relationships do not have a clear association with fragmentation per se but mainly with overall habitat loss [@Simberloff1982].
Aiming to design better management practices, this led to a nearly 50-year Single Large vs Several Small (SLOSS) debate. While some authors argue that habitat loss is the most overwhelming effect on species loss, for others the complex factors arising from fragmentation further enhance species extinctions [@Fahrig2017;@Fletcher2018].

The SLOSS debate has been an important academic and intellectual discussion in conservation biology. However, in practical implementation, numerous conservation agencies persist in prioritizing the safeguarding of extensive, contiguous habitat areas, often neglecting the protection of smaller patches of natural habitat [@Armsworth2018]. As an example, the current emphasis on 'rewilding' in Europe is geared towards the conservation and restoration of contiguous natural habitats, with a specified 'core area' of at least 10,000 hectares [@EuroparcFederation2013]. Similarly, conservation efforts for large water bodies take precedence, leaving small ponds or pondscapes largely excluded from priority considerations [@Hill2018].

While the overall habitat loss is widely accepted as a process leading to long-term biodiversity loss [@Keil2015;@Hodgson2011], the proposition that single large reserves retain more species than several small reserves of the same area (SL > SS) remain at a similar hierarchical state, even after several criticisms (reviewed in  @Fahrig2017). Empirical observational and experimental data shows the predominance of the opposite pattern (SS > SL) ever since the 80s and this fact has been reinforced to this day [@Simberloff1982; @Fahrig2020; @Hammill2020]. The actual processes driving it remain elusive and it is probably multi-causal. In that scenario, model simulations could produce useful hypotheses to better understand the processes leading to non-intuitive patterns of biodiversity loss.

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics. [cita?] However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

By examining spatially explicit models with a fixed level of habitat destruction alongside neutral and competitive community dynamics, encompassing diverse dispersal-related traits and various geometries of habitat loss, we aim to align the model parameters with the recently proposed SLOSS cube hypothesis  [@Fahrig2021]. This hypothesis serves as a comprehensive framework summarizing existing SLOSS predictions and empirical findings. According to the SLOSS cube hypothesis, the condition where SL > SS will prevail is contingent upon the simultaneous occurrence of the following conditions: low between-patch movement rate, minimal impact of spreading-of-risk on population dynamics, and low species spatial clumping [@Fahrig2021].

This work aims to explore spatial multi-species models to verify the conditions when SS > SL pattern holds and determine the causes of it. In particular, we simulate species retention after different fragmentation scenarios of the same overall habitat loss, therefore accounting for the interacting effects of habitat loss and fragmentation.

@@

We developed a spatially explicit multi-species model to explore the relationship between habitat loss and species retention at the regional metapopulation level. In this model, space is discretized into a grid and only one individual of one species can occupy a particular position. Each position represents an area fixed by the investigator to mimic the real system. Time is continuous, so the update of the model is asynchronous. Sites are selected at random and to perform one complete time interval $J$ sites have to be updated, where $J$ is the size of the grid.

We use absorbing boundary conditions, which means that if an individual tries to disperse outside the grid it dies. The size of the community is given by *J = dimX* x *dimY*, where *dimX* and *dimY* are the dimensions of the grid. Thus *J* is the maximum number of individuals in the simulated area. As in a classical neutral model, there is a metacommunity, i.e. a regional species pool from where species migrate [@Hubbell2001]. All individuals have the same parameters and each species is assigned an indicator number.

@@

The classical neutral variation is where propagules (e.g., seeds) disperse passively and cannot actively choose their establishment site. During reproduction events, individuals disperse their propagules blindly to surrounding patches. If a patch is already occupied, the propagule fails to establish, resulting in its death. This variation captures the random dispersal and colonization processes typical of many plant species.

The neutral variation with habitat selection is where individuals can actively select suitable habitat patches. In this variation, propagules can choose unoccupied patches, ensuring their survival and successful establishment. This modification reflects the influence of habitat availability and selection on species colonization and persistence. By incorporating habitat selection, this variation better captures the ecological processes and behaviors of species that are capable of actively seeking out suitable habitats for reproduction and survival, like most animals

The competitive hierarchical variation extends the neutral variation with an additional layer of complexity by incorporating competitive interactions among species. Each species has an indicator number that determines its competitive dominance level, and more dominant species can replace less dominant ones at a specified replacement rate. This variation captures the influence of species interactions and competitive exclusion on species composition and abundance. It is particularly relevant for understanding communities where competitive interactions play a significant role in shaping species distributions and community structure (e.g., hummingbirds).

@@

    Regular degradation: Habitat is removed so that habitable patches form a regular pattern of squares with sides given by a habitat-size parameter. As the percentage of degraded habitat is constant varying the habitat-size parameter results in either a few large habitat patches (SL) or several small (SS).

    Random block degradation: Habitat square patches of a fixed size (habitat-size parameter) are located at random throughout the landscape. The sizes are identical to those used for regular degradation simulations, but it results in habitat patches that have a minimum size corresponding to the habitat-size but a maximum size that is bigger because habitat patches can overlap between them producing a larger patch. Then, we characterize this with the maximum habitat patch size.

(2) With $migration-rate$ an individual of a species $i$ can migrate from the metacommunity. A random site on the border of the grid is selected and if it is empty and habitable the site is occupied[d].

@@

    where $d(x)$ is the probability that an individual disperses a distance $x$ from the parent. In all cases, we used $x_{min} = 1$. We use as parameter the mean $dispersal-distance$.

@@

    b. Neutral with Habitat Selection:  Propagules selectively target unoccupied patches for reproduction within the distance given by the dispersal kernel, ensuring their survival.

@@

For all simulations, we used a metacommunity of 100 species. The simulations start with the colonization of an empty habitat area. After 600 time steps fixed amount of 60% of habitat is subsequently subjected to destruction, employing regular or random block degradation to mimic habitat fragmentation due to habitat loss. <!-- why we don't use completely random degradation patterns -->

The parameters encompass a range of $growth-rates$ while holding the $death-rate$ fixed at 1. In this model, the relationship between these parameters, $\lambda = growth-rate / death-rate$, is the main determinant of the fate of an individual of a species. In similar non-spatial models, this relationship must exceed 1 for a species to persist. Thus we used a minimum $\lambda$ of 1.7, we also fixed the $migration-rate$ at 0.0001. We simulate each model variant using the parameters shown in Table 1. For the neutral classical version only, we also simulate random block habitat patterns. For each combination of parameters we ran 30 simulations up to 1500 time steps and recorded the number of species, the Shannon Diversity index [@Magurran2013] and the total number of individuals. For some representative combinations of parameters, we take snapshots of the model output showing the typical dynamics on time of the 3 recorded quantities and the rank abundance species distribution [@Newman2005; @May1975]

@@

The model was developed in Netlogo 6.3.0 and the model source code is available at Github <https://github.com/lsaravia/DynamicForestExtinction>, the Overview, Design concepts, Details (ODD), protocol for describing individual-based models is included within the model source code [@Grimm2010]. We performed the simulations and analysis using R version 4.3.1 [@RCoreTeam2023], the R source code to perform simulations and analysis is located at <https://github.com/EcoComplex/multiSLOSS>.

## Results[e]

In our neutral model simulations with a migration rate of $0.0001$ and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the $\lambda$ parameter (Figure 1). Species richness increased with $\lambda$, with the highest richness at smaller patch sizes for higher $\lambda$ values. Conversely, a $\lambda$ of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. In these cases only migrants survive in the smallest patches, and no local populations are established (Figures S13 & S14).
The interplay of dispersal distances, habitat sizes, and $\lambda$ impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low $\lambda$ and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

Our research contributes to the ongoing SLOSS debate in conservation biology, offering new insights based on our model findings. Consistent with previous studies, the results from our simulations primarily support the SS > SL pattern for species retention, as observed across various model scenarios [@Fahrig2020]. This trend underlines the greater efficacy of multiple smaller habitats in preserving species diversity compared to a single large habitat under equal conditions of habitat loss. A critical factor in this dynamic is the balance between reproduction and extinction rates ($\lambda$ parameter) and dispersal relative to the size of the habitat patches, influencing their role as sources of propagules. Notably, the SL > SS scenario emerges only when dispersal limitations prevent patch connectivity, and patch sizes are above a certain threshold to prevent local extinctions, aligning with Rybicki's [-@Rybicki2020] findings.

In neutral models one species by chance tends to dominate a patch [@Hubbell2001], so in fragmented landscapes, once the species increments its density it is unlikely that another species replaces it so with more patches the probability of having more species is augmented. This effect is enhanced when individuals can select the habitat, because of the increased survival of the propagules, thus fragmentation has a positive effect on the number of species. 
For competitive communities, there is a higher probability that different species dominate different smaller patches, compared with a bigger patch. Thus, again fragmentation has a positive effect on the number of species, this effect is called stochastic divergence [@Riva2023]. In smaller patches there is an increase in extinction risk because they support smaller populations [@Laurance2002;@Shafer1995], however as patches become smaller they also become more numerous, reducing the probability of extinction over the entire set of patches due to spreading the risk effect [@Crowley1977]. The combination of these two effects, the stochastic divergence and the risk effect, explain the highest number of species with patch size over the threshold of species survival.

@@

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

@@

The novel aspect of our model is that it considers communities that are assembled from a metacommunity in which individuals migrate to an empty habitat, and then after the number of species is stabilized the habitat is fragmented. This could be a more realistic scenario than if all species are present with initially random patterns at an already fragmented habitat [@Rybicki2020]. This could be more important the more heterogeneous was species spatial distribution pattern before fragmentation, as the cube hypothesis states.

In terms of the cube hypothesis the between patch movement is influenced by the dispersal distance in relation to patch size, as the landscape size and the percentage of habitat destruction are constant, the between-patch distance is higher for bigger patch sizes except for the bigest size that is only one patch, also the number of patches is higher for smaller patch sizes thus between patch movement will be higher. Dispersal distance also increase between patch movement, but for higher dispersal distances the sizes of the patches are less important and the relationship tends to be SS = SL.   The spreading of risk is influenced by the $\lambda$, for smaller lambda spreading the risk would be more important, but to fully incorporate the spreading the risk effect the model should include disturbances. Very low dispersal distance favours still favours combined with low $\lambda$ favours SL > SS up to a threshold after wich the patch size is not important. The species spatial clumping is influenced by the dispersal distance, higher dispersal distance produce less clumpines but this is cofounded with patch movement, so the effect is difficult to disentagle. Thus the cube hypothesis is generally compatible with the results of our model.

Our findings support Fahrig's [-@Fahrig2021] conclusion that conditions favoring SL > SS are limited and infrequently observed in nature. Under conditions of adequate patch size and dispersal capabilities, the SS scenario typically retains more species. This underscores the ecological significance of smaller habitat patches in maintaining biodiversity and landscape connectivity [@Yan2021; @Manning2006; @Shafer1995]. Our results thus challenge the conventional preference for larger reserves in conservation planning and support the paradigm shift in conservation towards considering sets of many small patches generally having higher biodiversity than sets of few large patches [@Riva2023].

In conclusion, we believe our study adds valuable perspectives to the SLOSS debate, illustrating the complexities of habitat fragmentation and species conservation. While our model has inherent limitations, it provides a useful framework for understanding key ecological processes. Future research should aim to extend these findings to more diverse ecological settings and incorporate additional factors influencing biodiversity in fragmented landscapes.

@@

[a]esta frase no me convence en general pero me parece que algo de esto es necesario para no saltar tanto desde lo que venimos hablando a el "here we..."
[b]Lo que trato de decir aca es que consideramos modelos de comunidades de distintos tipos, neutral y competitiva, con distintos traits de dispersion y distintos patrones de habitat loss.
[c]Cambié bastante la frase para agregar la discusion de la cube hypothesis.
[d]Hay 2 procesos migracion de la metacomunidad y dispersion local
[e]matendría en los plots el orden y nombres que están en el texto:
1) Neutral
2) Habitat Selection
3) Competitive Hierarchical
